\documentclass[notes,tikz,class=cs341]{agony}
% \includeonly{}
\graphicspath{{figures/}}
\usetikzlibrary{graphs,graphs.standard,graphdrawing,quotes,matrix}
\usegdlibrary{circular,force,trees,layered}
\makeatletter
\let\c@algorithm\relax
\@counteralias{algorithm}{theorem}
\makeatother

\title{CS 341 Spring 2023: Lecture Notes}
\begin{document}
\renewcommand{\contentsname}{CS 341 Spring 2023:\\{\huge Lecture Notes}}
\thispagestyle{firstpage}
\tableofcontents

Lecture notes taken, unless otherwise specified,
by myself during section 001 of the Spring 2023 offering of CS 350,
taught by Armin Jamshidpey.

\begin{multicols*}{2}
  \listoflecture
\end{multicols*}

\include{notes/01intro}
\include{notes/02divide}
\include{notes/03graph}
\include{notes/04greedy}
\include{notes/05dp}
\include{notes/06complexity}

\chapter{Final Review}
\lecture{(08/01)}

\begin{example}
  Suppose we want to schedule people be on call between time $S$ and $T$.
  If each of the $n$ people are available from $s_i$ to $t_i$,
  give a greedy algorithm to assign the minimum number of people.

  Assume that the input is already sorted by start time $S \leq s_1 \leq \dotsb \leq s_n \leq T$.
\end{example}
\begin{sol}
  Since we know we must start at $S$,
  consider all the intervals that start at $S$.
  Then, pick the one with the latest end time.

  For each subsequent selection, consider all the intervals with start times
  before the last chosen end time and then select the one with the latest end time.

  \begin{algorithm}[H]
    \caption{\Call{GreedyScheduleAssign}{$S,T,[s_1,t_1],\dotsc,[s_n,t_n]$}}
    \begin{algorithmic}[1]
      \State $O \gets \varnothing$
      \State $s \gets S$, $o \gets 1$
      \For{$i = 1,\dotsc,n$}
      \If{$s_i \leq s$}
      \If{$t_i \geq t_o$}
      \State $o \gets i$
      \EndIf
      \Else
      \State $O \gets O \cup \{o\}$
      \State $s \gets t_o$
      \State $o \gets i$
      \EndIf
      \EndFor
    \end{algorithmic}
  \end{algorithm}

  This runs in $O(n)$ time.
\end{sol}

\begin{example}
  An $i \times j$ rectangle is worth $P[i,j]$.
  Given an $n \times m$ rectangle, give a dynamic programming algorithm
  to find the optimal way to cut the rectangle into smaller rectangles.
\end{example}
\begin{sol}
  Let $M[i,j]$ be the optimal value of a rectangle after considering subdivisions.
  Then, $M[i,j]$ is either:
  \begin{itemize}
    \item $P[i,j]$, the value without cutting;
    \item $\displaystyle M_V(i,j) = \max\bigcup_{\mathclap{1 \leq k \leq i}}\{ M[k,j], M[i-k,j] \}$, the maximum value of a vertical cut; or
    \item $\displaystyle M_H(i,j) = \max\bigcup_{\mathclap{1 \leq k \leq j}}\{ M[i,k], M[i,j-k] \}$, the maximum value of a horizontal cut.
  \end{itemize}
  Finally, we just iterate.
  Each of the $O(nm)$ iterations takes $O(n+m)$ time, 
  so we have $O(n^2m + nm^2)$.
\end{sol}

\begin{example}
  Consider the problem $\vv{ModifiedSTPath}$:
  given an edge-weighted directed graph, is there a simple $s,t$-path
  (i.e., with no repeated vertices) with total weight at most $k$.
  What complexity class is $\vv{ModifiedSTPath}$ in?
\end{example}
\begin{sol}
  We can find shortest paths in polynomial time,
  so find a shortest path via Dijkstra and check if its weight is at most $k$.
  This means $\vv{ModifiedSTPath} \in \P$.
  Therefore, it is also in \NP and \coNP.
\end{sol}

\begin{example}
  Show that $\vv{ModifiedSTPath}$ is \NP-complete.
\end{example}
\begin{prf}
  We will show that $\vv{DirectedHamiltonianPath} \pleq \vv{ModifiedSTPath}$.

  Suppose we have a directed graph $G = (V,E)$ as input to $\vv{DirectedHamiltonianPath}$.
  Notice that a Hamiltonian path is just a simple path with $\abs{V}-1$ edges.

  If we give every edge in $G$ a weight of $-1$,
  then we can just count the edges that a path traverses.
  Let $G' = (V \cup \{s,t\}, E \cup \{sv, tv : v \in V\})$ and $w(e) = -1$.
  Then, we can call \Call{ModifiedSTPath}{$G', w, s, t, -(\abs{V}+1)$}
  which will respond ``yes'' exactly when there is a path of maximum length
  $\abs{V}+1$ and ``no'' otherwise.

  Therefore, $\vv{DirectedHamiltonianPath} \pleq \vv{ModifiedSTPath}$,
  which means $\vv{ModifiedSTPath}$ is \NP-complete, as desired.
\end{prf}

\phantomsection\addcontentsline{toc}{chapter}{Back Matter}

\renewcommand{\listtheoremname}{List of Problems}
\phantomsection\addcontentsline{toc}{section}{\listtheoremname}
\listoftheorems[ignoreall,onlynamed={problem}]

\renewcommand{\listtheoremname}{List of Named Results}
\phantomsection\addcontentsline{toc}{section}{\listtheoremname}
\listoftheorems[ignoreall,onlynamed={theorem,lemma,corollary}]

\printindex

\end{document}